\documentclass[a4paper, 11pt]{article}

\usepackage{xcolor}

\newcommand{\red}{\textcolor{red}}
\newcommand{\orange}{\textcolor{orange}}
\newcommand{\gray}{\textcolor{gray}}
\newcommand{\green}{\textcolor{green}}
\newcommand{\blue}{\textcolor{blue}}

\newcommand{\new}{\red{\emph{NEW!}} }
\newcommand{\maybe}{\orange{\emph{NEW?}} }

\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}
\usepackage{booktabs,hyperref}
\usepackage[longnamesfirst]{natbib}
\usepackage{array}

\begin{document}

\noindent {\LARGE Macroeconomics and Inequality}

\medskip \noindent {\large MSc Economics, University of Vienna, Summer Term 2026}

\medskip \noindent {{\textbf{Lecturer:} Dr. Fabian Greimel, \url{https://www.greimel.eu}}}

\medskip \noindent
This course provides a macroeconomic perspective on inequality, housing markets, and related structural issues. % such as demographic change and the climate transition. 
A central theme is that heterogeneity across households (differences in income, wealth, preferences, and investment opportunities) shapes aggregate outcomes and policy responses.

We develop tools and models that allow students to analyse and simulate heterogeneous-agent economies, building on the Bewley–Huggett–Aiyagari tradition and its modern extensions. The course blends theory, computation, and empirical context.

\medskip
\noindent\textbf{Core elements}
\begin{enumerate}
  \item \textbf{Methods:} Markov chains \& dynamic programming
  \item \textbf{Models:} Lifecycle and overlapping generations models, heterogeneous-agent models
  \item \textbf{Applications:} Wealth inequality, wealth taxation, social comparisons in the housng market, inequality and the response to macroeconomic shocks
  \item \textbf{Data:} Distributional National Accounts (DINA), the Survey of Consumer Finances (SCF), and the Survey of Consumer Expenditures (CEX)
\end{enumerate}

\medskip
The goal is to develop a sound command of modern macroeconomic models with heterogeneity—sufficient to understand current research and to implement numerical applications in the Julia programming language.





%\subsection*{Course Topics}
%
%\begin{tabular}{@{}lp{10cm}@{}}
%\toprule
%\textbf{Topic} & \textbf{Description} \\
%\midrule
%Lifecycle and OLG Models & $J$-period lifecycle, overlapping generations, infinite horizon \\
%Income and Borrowing & Income risk and borrowing constraints \\
%Wealth Distribution & Modeling wealth inequality and wealth taxes \\
%Welfare & Measuring social welfare for policy evaluation \\
%Housing & Housing market dynamics and their macroeconomic implications \\
%Microeconomic Heterogeneity & Impact on macroeconomic shocks and growth \\
%Markups and Inequality & Market power, price markups, and their effect on inequality \\
%\bottomrule
%\end{tabular}


\subsection*{Session by session}

\begin{tabular}{@{}c>{\raggedright\arraybackslash}p{8cm}>{\raggedright\arraybackslash}p{6.5cm}@{}}
\toprule
  & Topic & Reading \\
\midrule
1 & Introduction; history; two-period OLG & \citet{jappelli2017economics} \\
2 & $J$-period OLG &  \citet{jappelli2017economics} \\
3 & Dynamic programming: Value Function Iteration \\
4 & Income risk and borrowing constraints & \citet{jappelli2017economics} \\
5 & \new Income processes and Markov chains \\
6 & \new Dynamic programming: Endogenous Grid Method \\
7 & Lifetime uncertainty & \citet{jappelli2017economics} \\
8 & Bequests & \citet{jappelli2017economics}  \\ \midrule
9 & Tutorial &  \\
10 & Midterm exam \\ \midrule
11 & $\infty$-lived household as a Markov chain (I) \\
12 & $\infty$-lived household as a Markov chain (II) \\
13 & Modelling the wealth distribution \\
14 & Wealth inequality & \citet{hubmer2021sources} \\
15 & Housing &  \\
16 & \new Housing markets and wealth inequality & \citet{huber2025rental}  \\
17 & \new Measuring social welfare \& policy evaluation \\
18 & \new Entrepreneurs \\
19 & Wealth taxation (I) & \citet{guvenen2023use} \\
20 & Wealth taxation (II) & \citet{boar2023income-or-wealth} \\

21 & Social comparisons & \citet{drechsel2025falling-behind} \\
22 & \new Inequality and macroeconomic shocks & \citet{kaplan2018microeconomic} \\ \midrule
23 & Tutorial \\
24 & Final exam \\
\bottomrule
\end{tabular}

\subsection*{Note for myself: Building up to the Aiyagari model}

\begin{enumerate}
  \item Two period model without risk
  \item J-period model without risk
  \item Markov Chains
  \item Income process
  \item Two period model with risk
  \item exact (?), on an asset grid (== non-stochastic simulation)
\end{enumerate}

\subsection*{Bonus topic for the future: Markups and inequality}

\begin{itemize}
  \item Background I: \citet[JEP]{syverson2019macroeconomics}
  \item Background II \citet[QJE]{deloecker2020market-power}
  \item \url{https://cepr.org/voxeu/columns/price-power-why-rising-markups-hurt-innovation-and-widen-inequality}
  \item \citet[R\&R JEEA]{auray2026unpacking}
  \item \citet[R\&R REStud]{boar2024markups}
  \item Connection to HANK? \citet{auray2026unpacking} seem to be using the same specification of monopolistic competition.
\end{itemize}


\subsection*{Assessments criteria}

\begin{tabular}{llll}
\toprule
Weight & Criterion & Notes & \\
\midrule
30\% & 10 Assignments  & including computer assignments \& reading assignments \\
30\% & Midterm exam & in-class, closed-book \\
30\% & Final exam   & in-class, closed-book \\ 
10\% & Active participation & in-class engagement \\
\bottomrule
\end{tabular}

\begin{itemize}
  \item \emph{Computer assignments} are using the Julia programming language. These are meant to be simple, enlightening and fun!
  \item \emph{Reading assignments} will cover some of the papers listed in the references below.
  
  \item The tutorials will cover exercises in the style of the \emph{exams}. I will provide a \emph{sample exam} for both the midterm and the final exam. 
  
  \item \emph{Participation points} are awarded for active participation: questions, answers, comments. For full participation points, a student must actively participate in 15 sessions (other than the exams). The exact criteria will be discussed in the first session.
\end{itemize}


\subsection*{Grading scheme}

\begin{tabular}{ll}
\toprule
Grade & Points Range \\
\midrule
1 & [89, 100] \\
2 & [76, 89) \\
3 & [63, 76) \\
4 & [50, 63) \\
5 & [0, 50) \\
\bottomrule
\end{tabular}


\bibliographystyle{ecta}
\bibliography{../inequality.bib}

\end{document}
